### 2017-09-29 abstract.tex
completed abstract review, now working on introduction

% 
The validation of ground motion synthetics has received increased attention over the last few years due to the advances in physics-based deterministic and hybrid simulation methods. Unlike for low frequency simulations $(f \le 0.5 Hz)$, for which it has become reasonable to expect a good match between synthetics and data, in the case of high-frequency simulations $(1 Hz \le f)$ it is not possible to match results on a wiggle-by-wiggle basis. This is mostly due to the various complexities and uncertainties involved in earthquake ground motion modeling. Therefore, in order to compare synthetics with data we turn to different time series metrics, which are used as a means to characterize how the synthetics match the data on qualitative and statistical sense. In general, these metrics provide GOF scores that measure the level of similarity in the time and frequency domains. It is common for these scores to be scaled from 0 to 10, with 10 representing a perfect match. Although using individual metrics for particular applications is considered more adequate, there is no consensus or a unified method to classify the comparison between a set of synthetic and recorded seismograms when the various metrics offer different scores. We study the relationship among these metrics through 
semi-supervised and supervised learning approaches in order to labeling and classifying data, respectively. In particular we use constrained \kmeans{} clustering method, in which we define 4 hypothetical stations with scores 3, 5, 7, and 9 for all metrics. We put these stations in the category of cannot-link constraints. We generate the dataset through the validation of the results from a deterministic (physics-based) ground motion simulation for a moderate magnitude earthquake in the greater Los Angeles basin using three velocity models. The maximum frequency of the simulation is 4 Hz. The dataset involves over 300 stations and 11 metrics, or features, as they are understood in the machine learning lingo, where the metrics form a multi-dimensional space. We address the high-dimensional feature effects with a subspace-clustering analysis using 2,3, and 4 dimensional spaces. We label the dataset in each subspace as poor, fair, good, and excellent, where, hypothetical stations with score of 3,5,7, and 9 belongs them, respectively. We assign the final label based on the mode of subspaces for dataset.  Once the dataset has been labeled, we develop two decision trees using C5.0 algorithm to develop two prediction models to classify a simulation in four groups (poor, fair, good, and excellent) based on a reduced number of metrics. We address the unbalanced dataset issue through oversampling method and analyze the precision, recall and F1-score of the classifiers. The prediction models represent a simple, yet effective model to accurately classify the accuracy of the simulation process based on application-independent approach.

>>>>

% 
It has become common practice to validate ground motion simulation based on a variety of time- and frequency-domain metrics known in seismology and engineering and scaled to quantify the level of agreement between synthetics and data or other reference solutions. There is, however, no agreement about the importance or weight that it ought to be given to each individual metric. This leads to their selection often being subjective, either based on intended applications or personal preferences. In turn, this makes it difficult for simulators to identify modeling improvement needs, which could be easier if they could focused on a reduced number of alternative metrics. We present an analysis of ground motion validation metrics using semi-supervised and supervised learning techniques used to label and classify  goodness-of-fit (GOF) results, with the objective prioritizing and narrowing the choice of metrics available and commonly used in ground motion simulation validation. In particular, we study the relationships that exist between eleven different metrics when understood as part of a multi-dimensional space using a constrained \kmeans{} method. We conduct a subspace clustering analysis to address the high-dimensional effects and this allows us to label the results from a large validation dataset into four categories (poor, fair, good, and excellent). We then develop two decision trees using a C5.0 algorithm. These decision trees help narrow the number of metrics leading to a validation prediction into the four categories. These decision trees can be understood as rapid predictors of the quality of a simulation, or as a data-informed classifiers that can help prioritize validation metrics. Our analysis indicates that among the eleven metrics considered here, the acceleration response spectra and total energy of velocity are the most dominant metrics ones, followed by the peak ground response in terms of velocity and acceleration.

% OLD NAEEM VERSION
% 
% The validation of ground motion synthetics has received increased attention over the last few years due to the advances in physics-based deterministic and hybrid simulation methods. Unlike for low frequency simulations $(f \le 0.5 Hz)$, for which it has become reasonable to expect a good match between synthetics and data, in the case of high-frequency simulations $(1 Hz \le f)$ it is not possible to match results on a wiggle-by-wiggle basis. This is mostly due to the various complexities and uncertainties involved in earthquake ground motion modeling. Therefore, in order to compare synthetics with data we turn to different time series metrics, which are used as means to characterize how the synthetics match the data on qualitative and statistical sense. In general, these metrics provide GOF scores that measure the level of similarity in the time and frequency domains. It is common for these scores to be scaled from 0 to 10, with 10 representing a perfect match. Although using individual metrics for particular applications is considered more adequate, there is no consensus or a unified method to classify the comparison between a set of synthetic and recorded seismograms when the various metrics offer different scores. We study the relationship among these metrics through semi-supervised and supervised learning approaches in order to labeling and classifying data, respectively. In particular, we use constrained \kmeans{} clustering method, in which we define 4 hypothetical stations with scores 3, 5, 7, and 9 for all metrics. We put these stations in the category of cannot-link constraints. We generate the dataset through the validation of the results from a deterministic (physics-based) ground motion simulation for a moderate magnitude earthquake in the greater Los Angeles basin using three velocity models. The maximum frequency of the simulation is 4 Hz. The dataset involves over 300 stations and 11 metrics, or features, as they are understood in the machine learning lingo, where the metrics form a multi-dimensional space. We address the high-dimensional feature effects with a subspace-clustering analysis using 2,3, and 4 dimensional spaces. We label the dataset in each subspace as poor, fair, good, and excellent, where, hypothetical stations with score of 3,5,7, and 9 belong them, respectively. We assign the final label based on the mode of subspaces for each station.  Once the dataset has been labeled, we develop two decision trees using C5.0 algorithm as prediction models to classify a simulation in four groups (poor, fair, good, and excellent) based on a reduced number of metrics. We address the unbalanced dataset issue through oversampling method and analyze the precision, recall and F1-score of the classifiers. The prediction models represent a simple, yet effective model to accurately classify the accuracy of the simulation process based on application-independent approach. We analyze the different combinations of simulation accuracy classes for each components and present a guideline for making decision on accuracy level of simulation independent of components. We apply the proposed method on 2008 Chino Hills earthquake data and synthetic and present the results.

### 2017-09-29 introduction.tex
completed abstract review, now working on introduction

\section{Introduction}

The validation of ground motion synthetics has received increased attention over the last few years due to advances in physics based deterministic and hybrid simulation methods. Unless in very low frequencies (\fleq{0.5}), due to limitations and uncertainties in the ground motion simulation and velocity models, it is not possible to match the simulation and synthetics wiggle by wiggle. Therefor, in order to compare synthetic seismogram with recorded data as a complex time series, different metrics are defined. These metrics are means to characterize the how well the synthetic matches the statistical characteristic of observed data. In general, these metrics compute the goodness-of-fit (GOF) between data and synthetic in time and frequency domains. \citet{Anderson_2004_Proc} proposed 10 metrics to be used to characterized the GOF in different frequency bands. \citet{Kristekova_2006_BSSA,Kristekova_2009_GJI} developed envelope and phase misfit criteria as well as GOF metrics for realistic scenarios. \citet{Olsen_2010_SRL} presented an alternative GOF metrics for broadband synthetics with relatively high resolution of the small misfits, and \citet{Taborda_2013_BSSA} added modification to \citet{Anderson_2004_Proc} metrics.\\
Recently, due to advances in computational resources, high frequency 3D ground motion simulation on a regional scale were conducted in different regions using different methods (give some references.) The accuracy of simulation with observed data are measured using these GOF scores. However, researchers use these metrics in different perspectives which is highly dependent on their application. For example, for structural engineers who are interested in designing tall buildings, among other metrics, displacement could be an important factor. However, for rigid and semi-rigid structures like dams and nuclear power plants, acceleration could be more important. In general, for all purposes, the duration of strong ground morion could be an important factor. Analysis of the strong ground motion simulation for a specific application based on individual metrics is an acceptable practice, however, validating the simulation process only with individual score is not a valid approach. Our previous studies show that metrics could have completely different scores for a pair of synthetic and data \citep[e.g.,][]{Taborda_2016_GJI}. \citet{Taborda_2013_BSSA} defined a metric from a combination of  \citet{Anderson_2004_Proc} scores representing a general overview of GOF. However, the metric is a linear combination and somehow is biased by the authors preferences about the metrics. \\
Physics-based ground motion simulation community push the maximum frequency  of the simulation to higher values (\fmax{}$\geq 5 ~ Hz$) and there is a high demand for a metric to uniformly validate the simulations (refer to the high-f project). Although \citet{Anderson_2004_Proc}, defined different level for classifying the stations as poor, fair, good, and excellent based on individual metrics, the question becomes what if there is a station with completely different scores for different metrics. This research question motivated us to approach the problem from different perspective. In other words, if we assume we know the accuracy of the simulation process (let's say we classified it as good) what is the relationship between different GOF scores for that pair of stations. However, as far as the authors knowledge, there is no metric to be able to classify a pair of stations based on GOF scores as an estimation of the overall accuracy of the simulation. Therefore, as a first step, if we could label the pair of station for simulation accuracy we can discuss the relationships between GOF scores. To do this we need a good amount of dataset where calculated the GOF scores for a good amount of stations. \\  
 \citet{Taborda_2014_BSSA} conducted a comprehensive study to analyze the functionality of different velocity models in the Los Angeles basin. They used 2008 Chino Hills earthquakes based on 3 velocity models and conducted simulations with maximum frequency of \fmaxeq{4}. They compared the accuracy of the velocity models through using the GOF scores. The study generated an invaluable synthetic and data  GOF scores dataset.\\
The first and the most common method to categorizing multidimensional unlabeled data is clustering and specifically the distance based methods like \kmeans{} \citep{Macqueen_1967_Proc}. However, there is a drawback in this method. If the data is not explicitly distinguishable the final clusters are highly dependent on initial starting points of the algorithm. Therefor, for the use of stations with 11 dimensions, the ordinary \kmeans{} approach results in fairly different clusters after each analysis. The main reason for this issue is the algorithm is solely looking for data which are close together in terms of Euclidian distance. However, in our case we have one more extra information. Hypothetically, if we have a pair of synthetic and data with GOF score equal 3 for all metrics, according to \citet{Anderson_2004_Proc} we can classify the station as poor with a strong confidence. This is also true for GOF score of 5,7, and 9 for fair, good, and excellent classes, respectively. This background knowledge about the database could give a strong formation to the clustering process. \citet{Wagstaff_2001_Proc} defined a \kmeans{} clustering using background knowledge. Adding this feature to the algorithm improves the functionality and most importantly converge to the same clusters with any random initial cluster's centers. Using this approach, first we cluster stations then we statistically analyze each cluster data, based on these results we remove features with different trend from the analysis. Later on, in a supervised learning algorithm, we develop a classification algorithm based on their GOF scores.\\
In the following section,we present a summary of the validation metrics then we discuss the dataset. Later on we explain the methodology. We provide a brief summary of unsupervised learning which is ordinary and constrained clustering approaches. We also discuss the basics of decision tree algorithm as a supervised learning method. We discuss the clustering challenges and use subspace analysis. We develop two decision tree classifiers based on labeled data for future use in comparing data and synthetics. We conclude the study by applying the generated classification methods on \citet{Taborda_2014_BSSA} dataset.

>>>>

\introduction

The verification and validation of ground motion synthetics has received increasing attention in recent years due to advances in deterministic and non-deterministic physics-based earthquake ground motion simulation. Various methods or schemes have been proposed to evaluate, quantitatively, the misfit or goodness-of-fit (GOF) between simulation synthetics and recorded data or reference alternative solutions \citep{Anderson_2004_Proc, Kristekova_2006_BSSA, Kristekova_2009_GJI, Olsen_2010_SRL, Burks_BSSA_2014, Rezaeian_2015_BSSA}. Among these, there are methods better suited for verification purposes, because they target the comparisons at a waveform level \citep{Kristekova_2006_BSSA, Kristekova_2009_GJI}, whereas others evaluate the comparisons through proxies at the structural response level \citep{Burks_BSSA_2014}. The remaining ones, on the other hand, target specific characteristics in the seismograms themselves in both time and frequency, and are thus better suited for high-frequency or broadband simulations, where the expectation of matching seismograms at the waveform level are less relevant, unless dealing with low-frequency (\fmaxleq{1}). From this group, \citep{Anderson_2004_Proc, Olsen_2010_SRL} are preferred because they offer a simple scoring scheme to normalize the GOF scores associated with different metrics on a common scale.

Out of the various validation methods available, \citet{Anderson_2004_Proc} is perhaps the most widely used \citep[e.g.,]{Chaljub_2010_BSSA, Bielak_2010_GJI, Guidotti_2011_SRL, Maufroy_2015_BSSA}. We, in particular, have consistently relied on a modified version of this method for the validation of a series of simulations of the 2008 $M_w$ 5.4 Chino Hills earthquake, and for a larger suite of moderate-magnitude events in southern California \citep{Taborda_2013_BSSA, Taborda_2014_BSSA, Taborda_2016_GJI}. In essence, Anderson's method evaluates the similarity in terms of 10 metrics based on the cumulative strength of the motion (in terms of Arias intensity and absolute energy), the comparative evolution of the signals in time (through normalized Arias and energy integrals), the peak responses (in acceleration, velocity and displacement), the signals' frequency content (in the form of Fourier and response spectra), and their synchronization (through a cross correlation measure). While the strength

However, the existence of various alternative methods, including those inspired by or modified after Anderson's criteria, suggest the need to better understand what sort of parameters ought to be taken into account to appropriately classify the validity of a simulation.

Both \citet{Taborda_2013_BSSA} and \citet{Maufroy_2015_BSSA}, for instance, independently suggested the need to add

, and have more recently used it

% OLD NAEEM VERSION

% The validation of ground motion synthetics has received increased attention over the last few years due to the advances in physics based deterministic and hybrid simulation methods. Unless in very low frequencies (\fleq{0.5}), due to limitations and uncertainties in the ground motion simulation and velocity models, it is not possible to match the simulation and synthetics wiggle by wiggle. Therefor, in order to compare synthetic seismogram with recorded data as a complex time series, different metrics are defined. These metrics are means to characterize the how well the synthetic matches the statistical characteristic of observed data. In general, these metrics compute the goodness-of-fit (GOF) between data and synthetic in time and frequency domains. \citet{Anderson_2004_Proc} proposed 10 metrics to be used to characterize the GOF in different frequency bands. \citet{Kristekova_2006_BSSA,Kristekova_2009_GJI} developed envelope and phase misfit criteria as well as GOF metrics for realistic scenarios. \citet{Olsen_2010_SRL} presented an alternative GOF metrics for broadband synthetics with relatively high resolution of the small misfits, and \citet{Taborda_2013_BSSA} added modification to \citet{Anderson_2004_Proc} metrics.

% These metrics have been widely used in different studies. Recently, due to advances in computational resources, 3D ground motion simulation on a regional scale were conducted in different regions using different methods \citep[e.g.,][]{Taborda_2016_GJI,Komatitsch_2004_BSSA,Bielak_2010_GJI,Roten_2016_GJI}. The accuracy of simulation with observed data are measured using these GOF scores. However, researchers use these metrics in different perspectives which is highly dependent on their application. For example, for structural engineers who are interested in designing tall buildings, among other metrics, displacement could be an important factor. However, for rigid and semi-rigid structures like dams and nuclear power plants, acceleration could be more important criteria. In general, for all purposes, the duration of strong ground morion could be an important factor. Analysis of the strong ground motion simulation for a specific application based on individual metrics is an acceptable practice, however, validating the simulation process only with individual score is not a valid approach. Previous studies show that metrics could have completely different scores for a pair of synthetic and data \citep[e.g.,][]{Taborda_2016_GJI}. \citet{Taborda_2013_BSSA} defined a metric from a combination of  \citet{Anderson_2004_Proc} scores presenting a general overview of GOF. However, the metric is a linear combination and somehow is biased by the authors' preferences about the metrics.

% Physics-based ground motion simulation communities push the maximum frequency  of the simulation to higher values (\fmax{}$\geq 5 ~ Hz$) and there is a high demand for a metric to uniformly validate the simulations. Although \citet{Anderson_2004_Proc}, defined different level for classifying the stations as poor, fair, good, and excellent based on individual metrics, there is no consensus on the most important metrics, specially when they have significantly different GOF score. Therefore, the question "How one can estimate the accuracy of simulation based on all scores or which metrics can be good representatives of simulation accuracy on behalf of other metrics?" has not addressed. This research question motivated us to approach the problem from different perspective. In other words, if we assume we know the accuracy of the simulation process (let's say we classified it as good) what is the relationship between different GOF scores for that pair of stations or which metrics are the best representer in classifying the stations. However, as far as the authors knowledge, there is no metric to be able to classify a pair of stations based on GOF scores as an estimation of the overall accuracy of the simulation. Therefore, as a first step, if we label the pair of station for simulation accuracy we can discuss the relationships between GOF scores. To do this we need a dataset where involves GOF scores for a good amount of stations.

% \citet{Taborda_2014_BSSA} conducted a comprehensive study to analyze the functionality of different velocity models in the Los Angeles basin. They used 2008 Chino Hills earthquakes based on 3 velocity models and conducted simulations with maximum frequency of \fmaxeq{4}. They compared the accuracy of the velocity models through using the GOF scores. The study generated an invaluable synthetic and data  GOF scores dataset.\\
% The first and the most common method in categorizing multidimensional unlabeled data is clustering and specifically the distance based methods like \kmeans{} \citep{Macqueen_1967_Proc}. However, there is a drawback in this method. If the data is not explicitly distinguishable the final clusters are highly dependent on initial starting points of the algorithm. Therefor, for stations with 11 dimensions, the ordinary \kmeans{} approach results in fairly different clusters after each analysis. The main reason for this issue is the algorithm solely looks for data which are close together in terms of Euclidean distance. However, in our case we have one more extra information. Hypothetically, if we have a pair of synthetic and data with GOF score equal 3 for all metrics, according to \citet{Anderson_2004_Proc} we can classify the station as poor with a strong confidence. This is also true for GOF score of 5, 7, and 9 for fair, good, and excellent classes, respectively. This background knowledge about the database could give a strong formation to the clustering process. \citet{Wagstaff_2001_Proc} defined a \kmeans{} clustering using background knowledge. Adding this feature to the algorithm improves the functionality and most importantly converge to the same clusters with any random initial clusters' centers. Using this approach alongside the subspace analysis (more details are presented in the section.~\ref{clustering}) we cluster the stations, then in a supervised learning algorithm, we develop a classification algorithm based on their GOF scores. We tune the learning algorithm and present the functionality of each prediction model based on different metrics. It is possible to have different classes for different components of a pair if synthetic and data. We study the different combination of components' classes in different distances and propose a rough estimate on deciding about the final GOF class of a pair of synthetic and data which is application and component independent.\\
% In the following section,we present a summary of the validation metrics then we discuss the dataset. Later on we explain the methodology. We provide a brief summary of unsupervised  and semi-supervised learning which are ordinary and constrained clustering approaches, respectively. We also discuss the basics of decision tree algorithm as a supervised learning method. We discuss the clustering challenges and use subspace analysis in the clustering analysis section. We also discuss the imbalanced data effects and solutions in the classification model section. We develop two decision tree classifiers (prediction model) based on labeled data for future use in comparing data and synthetics. We propose an algorithm to give a rough estimate to have component independent prediction. We conclude the study by applying the generated classification methods on \citet{Taborda_2014_BSSA} dataset and present the spatial variation of the simulation goodness of fit classes on the Los Angles basin for three velocity models and three components in component dependent and for three velocity models in component independent approaches.
